Possible Inconsistency in Maximum Likelihood Calculation of Ancestral States

### 2017-04-28 deprecated.tex
add appendix; address most issues

@@

\begin{table}
\centering
\begin{tabular}{|l|ll|}
\multicolumn{3}{c}{$P(\ancestralStateColumn=\ancestralSplit_k|\alignmentColumn=\siteSplit_j,\tau_1,t)$}\\
\hline
& \multicolumn{2}{|c|}{$\ancestralSplit_k$}\\
    \hline
    $\siteSplit_j$    &$\emptyset$  00                              &$\{2\}$  01\\
    \hline
     $\emptyset$   0000&$(1+x)^2   (1+w)(1+y)^{2*}$          &$(1+x)^2   (1-w)(1-y)^2$\\
     $\{1\}$       1000&$(1+x)(1-x)(1+w)(1+y)^{2*}$          &$(1+x)(1-x)(1-w)(1-y)^2$\\
     $\{2\}$       0100&$(1+x)(1-x)(1+w)(1+y)^{2*}$          &$(1+x)(1-x)(1-w)(1-y)^2$\\
     $\{3\}$       0010&$(1+x)^2   (1+w)(1+y)(1-y)^*$        &$(1+x)^2   (1-w)(1+y)(1-y)$\\
     $\{1,2\}$     0011&$(1+x)(1-x)(1+w)(1+y)(1-y)^{\dagger}$&$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
     $\{1,3\}$     0101&$(1+x)^2   (1+w)(1-y)^{2\ddagger}$   &$(1+x)^2   (1-w)(1+y)^{2\ddagger}$\\
     $\{2,3\}$     1001&$(1+x)(1-x)(1+w)(1+y)(1-y)^{\dagger}$&$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
     $\{1,2,3\}$   0001&$(1+x)^2   (1+w)(1+y)(1-y)^*$        &$(1+x)^2   (1-w)(1+y)(1-y)$\\
    \hline
    \hline
    &$\{1\}$  10                           &$\{1,2\}$  11\\
    \hline
     $\emptyset$   0000&$(1-x)^2   (1-w)(1+y)^2$     &$(1-x)^2   (1+w)(1-y)^2$\\
     $\{1\}$       1000&$(1+x)(1-x)(1-w)(1+y)^2$     &$(1+x)(1-x)(1+w)(1-y)^2$\\
     $\{2\}$       0100&$(1+x)(1-x)(1-w)(1+y)^2$     &$(1+x)(1-x)(1+w)(1-y)^2$\\
     $\{3\}$       0010&$(1-x)^2   (1-w)(1+y)(1-y)$  &$(1-x)^2   (1+w)(1+y)(1-y)$\\
     $\{1,2\}$     0011&$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)^{\dagger}$\\
     $\{1,3\}$     0101&$(1-x)^2   (1-w)(1-y)^2$     &$(1-x)^2   (1+w)(1+y)^{2\ddagger}$\\
     $\{2,3\}$     1001&$(1+x)(1-x)(1-w)(1+y)(1-y)$  &$(1+x)(1-x)(1+w)(1+y)(1-y)^{\dagger}$\\
     $\{1,2,3\}$   0001&$(1-x)^2   (1-w)(1+y)(1-y)$  &$(1-x)^2   (1+w)(1+y)(1-y)$\\
    \hline
\multicolumn{3}{c}{$P(\ancestralStateColumn=\ancestralSplit_k|\alignmentColumn=\siteSplit_j,\tau_2,t)$}\\
\hline
& \multicolumn{2}{|c|}{$\ancestralSplit_k$}\\
    \hline
    $\siteSplit_j$    &$\emptyset$  00                              &$\{2\}$  01\\
    \hline
     $\emptyset$   0000&$(1+x)^2   (1+w)(1+y)^{2*}$           &$(1+x)(1-x)(1-w)(1+y)(1-y)$\\
     $\{1\}$       1000&$(1+x)(1-x)(1+w)(1+y)^{2\ddagger}$    &$(1-x)^2   (1-w)(1+y)(1-y)$\\
     $\{2\}$       0100&$(1+x)^2   (1+w)(1+y)(1-y)^{\ddagger}$&$(1+x)(1-x)(1-w)(1-y)^2$\\
     $\{3\}$       0010&$(1+x)(1-x)(1+w)(1+y)^{2\ddagger}$    &$(1+x)^2   (1-w)(1+y)(1-y)^{\ddagger}$\\
     $\{1,2\}$     0011&$(1+x)(1-x)(1+w)(1+y)(1-y)^{\ddagger}$&$(1+x)^2   (1-w)(1+y)^{2\ddagger}$\\
     $\{1,3\}$     0101&$(1+x)^2   (1+w)(1-y)^{2\ddagger}$    &$(1+x)(1-x)(1-w)(1+y)(1-y)^{\ddagger}$\\
     $\{2,3\}$     1001&$(1+x)(1-x)(1+w)(1+y)(1-y)^{\ddagger}$&$(1-x)^2   (1-w)(1+y)^{2\ddagger}$\\
     $\{1,2,3\}$   0001&$(1+x)^2   (1+w)(1+y)(1-y)^{\ddagger}$&$(1+x)(1-x)(1-w)(1+y)^{2\ddagger}$\\
    \hline
    \hline
    &$\{1\}$  10                           &$\{1,2\}$  11\\
    \hline
     $\emptyset$   0000&$(1+x)(1-x)(1-w)(1+y)(1-y)$             &$(1-x)^2   (1+w)(1-y)^2$\\
     $\{1\}$       1000&$(1+x)^2   (1-w)(1+y)(1-y)^{\ddagger}$  &$(1+x)(1-x)(1+w)(1-y)^2$\\
     $\{2\}$       0100&$(1+x)(1-x)(1-w)(1+y)^{2\ddagger}$      &$(1-x)^2   (1+w)(1+y)(1-y)$\\
     $\{3\}$       0010&$(1-x)^2   (1-w)(1+y)(1-y)$             &$(1+x)(1-x)(1+w)(1-y)^2$\\
     $\{1,2\}$     0011&$(1-x)^2   (1-w)(1-y)^2$                &$(1+x)(1-x)(1+w)(1+y)(1-y)^{\ddagger}$\\
     $\{1,3\}$     0101&$(1+x)(1-x)(1-w)(1+y)(1-y)^{\ddagger}$  &$(1-x)^2   (1+w)(1+y)^{2\ddagger}$\\
     $\{2,3\}$     1001&$(1+x)^2   (1-w)(1-y)^{2\ddagger}$      &$(1+x)(1-x)(1+w)(1+y)(1-y)^{\ddagger}$\\
     $\{1,2,3\}$   0001&$(1+x)(1-x)(1-w)(1-y)^2$                &$(1-x)^2   (1+w)(1+y)(1-y)$\\
\hline
\end{tabular}
\caption{Likelihood calculations for all site patterns and internal states of Farris topology.
Maxima determined row-wise (i.e., by site pattern).
All values multiplied by $1/32$.
Key: $^*$ unique maximum value corresponding to unique internal state; $^\dagger$ unique maximum value corresponding to multiple internal states; $^\ddagger$ multiple maximum values corresponding to multiple internal states.}
\label{tab:likelihoods}
\end{table}

>>>>

\paragraph{Examples}

See Table~\ref{tab:examples} for parameters, bounds and true likelihood values obtained via basin hopping.

Set $\{x, y\} = \{0.25, 0.25\}$.
For a Farris zone generating topology, we have
$$
L^{p,1}_{\tau_1,\tau_1}(\mathbf{a}_{xy}) \approx -2.17803560694157,
$$
$$
L^{p,2}_{\tau_1,\tau_1}(\mathbf{a}_{xy}) \approx -2.17803560694157
$$
and
$$
L^{p,3}_{\tau_1,\tau_1}(\mathbf{a}_{xy}) \approx -2.17803560694157.
$$
Computing the upper bound yields
$$
\max_{t,\fullAncestralSplitPartitions(\tau_1,t)} \ \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_1,t),\tau_1,t) \le -4.22274479428471
$$
and it remains to show that
$$
\max_{t,\fullAncestralSplitPartitions(\tau_2,t)} \ \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_2,t),\tau_2,t) \ge -4.22274479428471.
$$

For the same generating parameters,
$$
d_{xy} = 0.498046875, \ e_{xy} = 0.2431640625,
$$
resulting in
$$
L^{p,1}_{\tau_1,\tau_2}(d_{xy},e_{xy}) \approx -2.07968569223991
$$
and
$$
\shannonDivergence_{\tau_1,\{x,y,y\}}(\tau_2,\{1, 1-d_{xy}-2e_{xy}, 0\}) \approx -2.07748833719921.
$$
The lower bound in this case becomes
$$
\max_{t,\fullAncestralSplitPartitions(\tau_2,t)} \ \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_2,t),\tau_2,t) \ge -4.15717402943912
\ge -4.22274479428471 \ge \max_{t,\fullAncestralSplitPartitions(\tau_1,t)} \ \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_1,t),\tau_1,t)
$$
where site patterns are generated from the Farris zone topology.
Confirming, run \texttt{python}'s numerical optimization solver for $\{.25, .25\}$.
Since we are computing a constrained maximization, we can find a global maximum using basin hopping, and doing so yields
$$
\max_{t,\fullAncestralSplitPartitions(\tau_1,t)} \ \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_1,t),\tau_1,t) \approx -4.5101771294218045
$$
and
$$
\max_{t,\fullAncestralSplitPartitions(\tau_2,t)} \ \ell_{\tau_1,\{x,y,y\}}(\fullAncestralSplitPartitions(\tau_2,t),\tau_2,t) \approx -4.1544859851260281.
$$
It looks like the lower bound is tighter than the upper bound, so we may try focusing there.

@@
